---
title: "Boone"
author: "Sam Tolbert"
date: "2024-10-02"
output: 
  pdf_document:
    toc: true
output_dir: "Hydrology/Report"
---

```{r setep}
knitr::opts_chunk$set(echo = FALSE)

```

Initial Set-Up

```{r initialize project}

library(dplyr)
library(ggplot2)
library(lubridate)

Boone <- read.csv("Hydrology/Data/Raw/Boone_daily_precip_1980-present_HUC_050500010201_dayMet_split-dates-columns.csv", stringsAsFactors = TRUE)

Boone_Data<-Boone

# Rename the precipitation column to 'Precipitation in mm'
Boone_Processed <- Boone_Data %>%
  rename(Precipitation_mm = Area.Weighted.Mean.Precipitation..mm.per.day.)

# Ensure the 'Date' column is in date format
Boone_Processed <- Boone_Processed %>%
  mutate(Date = as.Date(Date))

```

**All code was produced in conversation with R Wizard GPT. Prompts included
in code where relevant.

----
The first analysis we can run is finding large scale trends within the
precipitation data. Using the HUC data, we found that a general increase in
annual precipitation. We find that from 1980 to 2016 there has been a steady
increase in the annual mean preciptation

```{r}

#First we isolate the means of each year:
Boone_Annual_Averages <- Boone_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(year) %>%
  summarize(annual_avg_precip = mean(Precipitation_mm, na.rm = TRUE))


# The we plot the annual averages using a bar plot with a linear regression line
ggplot(Boone_Annual_Averages, aes(x = year, y = annual_avg_precip)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Bar plot
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +  # Linear regression line
  labs(title = "Annual Average Precipitation (1980-2016) with Linear Trend",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal()

```



While this general trend is useful, in regard to infrastructure,
it is vital to know if this precipitation is evenly distributed 
throughout the year, is concentrated in larger, predictable hurricane systems
(Hurricane Season defined by the State of North Carolina as June 1-November 30)
or the less predictable, smaller frontal systems throughout the rest of the
year. We find that the rainest months occur during Hurricane season, with
July, September, and June being the rainiest months.

```{r, echo=TRUE}

#Average Monthly Precipitation

Boone_Monthly_Averages_AllYears <- Boone_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(month) %>%
  summarize(monthly_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

# Plotting the monthly averages (across all years) using a bar plot
ggplot(Boone_Monthly_Averages_AllYears, aes(x = factor(month), y = monthly_avg_precip)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Average Monthly Precipitation (1980-2016)",
       x = "Month",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_x_discrete(labels = month.name)  # Adding month names to the x-axis

```


Combining these two analyses we can see the monthly trends from 1980-2016, seeing
a steady increase in mean precipation, particularly in the summer months

```{r, echo=TRUE}

# Group by year and month, and calculate the mean precipitation for each month
Boone_Monthly_Averages <- Boone_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(year, month) %>%
  summarize(monthly_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

# Plotting the monthly averages using a bar plot
ggplot(Boone_Monthly_Averages, aes(x = factor(month), y = monthly_avg_precip, fill = factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Monthly Average Precipitation (1980-2016)",
       x = "Month",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_fill_viridis_d(name = "Year") +
  scale_x_discrete(labels = month.name)  # Adding month names to the x-axis



```


Seeing these trends, we now compare the mean precipiations of Hurricane Season
against Frontal systems.
```{r pressure, echo=TRUE}

#Creating Hurricane Season vs. Frontal Dataframe
Boone_Seasonal <- Boone_Processed %>%
  mutate(Season = case_when(
    (month >= 6 & month <= 11) ~ "Hurricane Season",  # June to November
    TRUE ~ "Frontal"  # December to May
  ))

#Filter data for the years 1980-2016
Boone_Seasonal <- Boone_Seasonal %>%
  filter(year >= 1980 & year <= 2016)

#Grouping by year and season, we calculate the average precipitation for each year and season
Boone_Seasonal_Averages <- Boone_Seasonal %>%
  group_by(year, Season) %>%
  summarize(avg_precip = mean(Precipitation_mm, na.rm = TRUE))
```


We express the results with a bar plot with separate linear regression lines for each season
```{r}

ggplot(Boone_Seasonal_Averages, aes(x = year, y = avg_precip, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +  # Bar plot for both seasons side-by-side
  geom_smooth(method = "lm", aes(color = Season), se = FALSE) +  # Add separate linear regression lines for each season
  labs(title = "Average Precipitation for Hurricane Season vs Frontal (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_fill_manual(values = c("Hurricane Season" = "skyblue", "Frontal" = "orange")) +
  scale_color_manual(values = c("Hurricane Season" = "blue", "Frontal" = "red"))
```

This graph makes it apparent that Hurricane Season has been steadily increasing
in terms of mean precipation in the last 36 years, while frontal precipation
has been staying flat. Combining this with our previous analysis, we can assume
that the majority of the annual precipitation increases are due to
increased intensities and rainfall of hurricane systems,
rather than frontal systems.

Here are those graphs again, side by side:

```{r}
ggplot(Boone_Seasonal_Averages, aes(x = year, y = avg_precip)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot for average precipitation
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear regression line
  facet_wrap(~ Season) +  # Facet by season ("Hurricane Season" and "Frontal")
  labs(title = "Seasonal Precipitation Trends (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal()
```

We can also see this incrasing dominance of Hurricane systems, especially
since the year 2000 in the followng Time Series Analysis:
```{r}
ggplot(Boone_Seasonal_Averages, aes(x = year, y = avg_precip, color = Season, group = Season)) +
  geom_line(size = 1) +
  labs(title = "Time Series of Precipitation (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_color_manual(values = c("Hurricane Season" = "blue", "Frontal" = "red"))

#these two grapsh show that if you give equal weight to pre 2000 post 2000, then the frontal season shifts from a flat trend to a rising trend

```


These increasing trends become even more apparent when we give equal weight
to pre-2000 and post-2000 date, filtering data to give a 16 year 
year window pre/post 2000.


```{r, echo=TRUE}
Boone_Seasonal_Averages_Modified <- Boone_Seasonal_Averages %>%
  filter(year >= 1984 & year <= 2016) #creating a data set from 1984-2016

# Plot with separate regression lines for each season (1984-2016)
ggplot(Boone_Seasonal_Averages_Modified, aes(x = year, y = avg_precip)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot for average precipitation
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear regression line
  facet_wrap(~ Season) +  # Facet by season ("Hurricane Season" and "Frontal")
  labs(title = "Seasonal Precipitation Trends (1984-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal()


```



This leads us to begin to want to track the temporal changes, not just the
seasonal. How have intensities altered intra season since date started being
collected?


```{r, Hazen Mutuate, echo=TRUE}


# Load necessary libraries
library(dplyr)
library(lubridate)

# Assuming the data is already loaded into Boone

# Create new columns for year if not already present
Boone_Data <- Boone_Data %>%
  mutate(Date = as.Date(Date))  # Ensure the Date column is in Date format

# Function to calculate Fa (Return Period) using the Hazen Equation
# n is the rank and y is the total number of observations
calc_fa <- function(n, y) {
  return(100 * (2 * n - 1) / (2 * y))
}

# Filter for data between 1980-01-01 and 2000-12-31 (daily resolution)
Boone_filtered_19802000 <- Boone_Data %>%
  filter(Date >= as.Date("1980-01-01") & Date <= as.Date("2000-12-31"))

# Rank the precipitation data for 1980-2000 in descending order (highest to lowest)
Boone_Hazen_Mutate_19802000 <- Boone_filtered_19802000 %>%
  arrange(desc(`Area.Weighted.Mean.Precipitation..mm.per.day.`)) %>%
  mutate(
    rank = row_number(),  # Assign ranks where 1 is the highest precipitation
    Fa = calc_fa(rank, n())  # Calculate Fa using Hazen Equation
  ) %>%
  mutate(likelihood_event = 100 / Fa)  # Add likelihood column

# Filter for data between 2000-01-01 and 2016-12-31 (daily resolution)
Boone_filtered_20002016 <- Boone_Data %>%
  filter(Date >= as.Date("2000-01-01") & Date <= as.Date("2016-12-31"))

# Rank the precipitation data for 2000-2016 in descending order (highest to lowest)
Boone_Hazen_Mutate_20002016 <- Boone_filtered_20002016 %>%
  arrange(desc(`Area.Weighted.Mean.Precipitation..mm.per.day.`)) %>%
  mutate(
    rank = row_number(),  # Assign ranks where 1 is the highest precipitation
    Fa = calc_fa(rank, n())  # Calculate Fa using Hazen Equation
  ) %>%
  mutate(likelihood_event = 100 / Fa)  # Add likelihood column





```


heatmap at the bottom because it looks cool but I don't think we ultimately need?
```{r}



Boone_Heatmap_Data <- Boone_Processed %>%
  group_by(year, month) %>%
  summarize(monthly_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

ggplot(Boone_Heatmap_Data, aes(x = year, y = factor(month), fill = monthly_avg_precip)) +
  geom_tile() +
  labs(title = "Heatmap of Monthly Average Precipitation (1980-2016)",
       x = "Year",
       y = "Month",
       fill = "Precipitation (mm)") +
  scale_fill_viridis_c() +
  scale_y_discrete(labels = month.name) +
  theme_minimal()