---
title: "Greensboro Report"
author: "Sam Tolbert"
date: "2024-10-02"
output: 
  pdf_document:
    toc: false  # Optional: Adds a table of contents
output_dir: "Hydrology/Report"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Raw Data see broad annual and monthly trends from 1980-2016

```{r}

library(dplyr)
library(ggplot2)
library(lubridate)

Greensboro <- read.csv("Hydrology/Data/Raw/Greensboro_daily_precip_1980-present_HUC_030300020105_dayMet_split-dates-columns.csv", stringsAsFactors = TRUE)

Greensboro_Data <- Greensboro

# Rename the precipitation column to 'Precipitation in mm'
Greensboro_Processed <- Greensboro_Data %>%
  rename(Precipitation_mm = Area.Weighted.Mean.Precipitation..mm.per.day.)

# Ensure the 'Date' column is in date format
Greensboro_Processed <- Greensboro_Processed %>%
  mutate(Date = as.Date(Date))

```

*All code was produced in conversation with R Wizard GPT. Prompts included in code where relevant.
----

The first analysis we can run is finding large scale trends within the precipitation data. Using the HUC data, we found a constant trend in annual precipitation from 1980-2016


```{r}


# First, we isolate the means of each year:
Greensboro_Annual_Averages <- Greensboro_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(year) %>%
  summarize(annual_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

# The we plot the annual averages using a bar plot with a linear regression line
ggplot(Greensboro_Annual_Averages, aes(x = year, y = annual_avg_precip)) +
  geom_bar(stat = "identity", fill = "skyblue") +  # Bar plot
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +  # Linear regression line
  labs(title = "Greensboro Annual Average Precipitation (1980-2016) with Linear Trend",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal()




```


While this general trend is useful, in regard to infrastructure, it is vital to know if this precipitation is evenly distributed throughout the year, is concentrated in larger, predictable hurricane systems (Hurricane Season defined by the State of North Carolina as June 1-November 30) or the less predictable, smaller frontal systems throughout the rest of the year. We find that the rainest months occur during Hurricane season, with July, September, and March being the rainiest months.

```{r}



#Average Monthly Precipitation

Greensboro_Monthly_Averages_AllYears <- Greensboro_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(month) %>%
  summarize(monthly_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

# Plotting the monthly averages (across all years) using a bar plot
ggplot(Greensboro_Monthly_Averages_AllYears, aes(x = factor(month), y = monthly_avg_precip)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Greensboro Average Monthly Precipitation (1980-2016)",
       x = "Month",
       y = "Average Precipitation (mm)") +
  theme_minimal() + theme(
    axis.text.x = element_text(angle=45, 
                               hjust=1))+
  scale_x_discrete(labels = month.name)  # Adding month names to the x-axis


```
Combining these two analyses we can see the monthly trends from 1980-2016, seeing a steady increase in mean precipation, particularly in the summer months.

```{r}



# Group by year and month, and calculate the mean precipitation for each month
Greensboro_Monthly_Averages <- Greensboro_Processed %>%
  filter(year >= 1980 & year <= 2016) %>%
  group_by(year, month) %>%
  summarize(monthly_avg_precip = mean(Precipitation_mm, na.rm = TRUE))

# Plotting the monthly averages using a bar plot
ggplot(Greensboro_Monthly_Averages, aes(x = factor(month), y = monthly_avg_precip, fill = factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Greensboro Monthly Average Precipitation (1980-2016)",
       x = "Month",
       y = "Average Precipitation (mm)") +
  theme_minimal() + 
  theme(
    axis.text.x = element_text(
      angle=45, 
      hjust=1))+
  scale_fill_viridis_d(name = "Year") +
  scale_x_discrete(labels = month.name)  # Adding month names to the x-axis




```
Seeing these trends, we now compare the mean precipiations of Hurricane Season against Frontal systems.
```{r}

#Creating Hurricane Season vs. Frontal Dataframe
Greensboro_Seasonal <- Greensboro_Processed %>%
  mutate(Season = case_when(
    (month >= 6 & month <= 11) ~ "Hurricane Season",  # June to November
    TRUE ~ "Frontal"  # December to May
  ))

#Filter data for the years 1980-2016
Greensboro_Seasonal <- Greensboro_Seasonal %>%
  filter(year >= 1980 & year <= 2016)

#Grouping by year and season, we calculate the average precipitation for each year and season
Greensboro_Seasonal_Averages <- Greensboro_Seasonal %>%
  group_by(year, Season) %>%
  summarize(avg_precip = mean(Precipitation_mm, na.rm = TRUE))

```

We express the results with a bar plot with separate linear regression lines for each season.

```{r}

ggplot(Greensboro_Seasonal_Averages, aes(x = year, y = avg_precip, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +  # Bar plot for both seasons side-by-side
  geom_smooth(method = "lm", aes(color = Season), se = FALSE) +  # Add separate linear regression lines for each season
  labs(title = "Greensboro Average Precipitation for Hurricane Season vs Frontal (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_fill_manual(values = c("Hurricane Season" = "skyblue", "Frontal" = "orange")) +
  scale_color_manual(values = c("Hurricane Season" = "blue", "Frontal" = "red"))



```


This graph makes it apparent that despite the constant trend in precipitation, Hurricane Season has been steadily increasing in terms of mean precipation in the last 36 years, while frontal precipation has been decreasing. Combining this with our previous analysis, we can assume that the majority of the despite constant trends in precipitation, hurricane season is becoming heavier due to increased intensities and rainfall of hurricane systems, rather than frontal systems.

Here are those graphs again, side by side:

```{r}
ggplot(Greensboro_Seasonal_Averages, aes(x = year, y = avg_precip)) +
  geom_bar(stat = "identity", fill = "steelblue") +  # Bar plot for average precipitation
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear regression line
  facet_wrap(~ Season) +  # Facet by season ("Hurricane Season" and "Frontal")
  labs(title = "Greensboro Seasonal Precipitation Trends (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal()


```

We can also see this increasing dominance of Hurricane systems, especially since the year 2000 in the following Time Series Analysis:



```{r}
ggplot(Greensboro_Seasonal_Averages, aes(x = year, y = avg_precip, color = Season, group = Season)) +
  geom_line(size = 1) +
  labs(title = "Greensboro Time Series of Precipitation (1980-2016)",
       x = "Year",
       y = "Average Precipitation (mm)") +
  theme_minimal() +
  scale_color_manual(values = c("Hurricane Season" = "blue", "Frontal" = "red"))

```




