---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Sam Tolbert"
date: "Fall 2024"
output: pdf_document
toc : true
lof : true
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message= FALSE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

getwd() #home/gues/EDA Fall 2024
library(tidyverse)
library(dplyr)
library(agricolae)
library(lubridate)
library(ggplot2)
library(here)

NTLakes<-read.csv(here( "Data", "Raw",
                         "NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                  stringsAsFactors = T
                  )

NTLakes$sampledate<-mdy(NTLakes$sampledate)



#2

SamTheme<- theme(
  text = element_text(family="Helvetica", size= 12),
  plot.title = element_text(size=16, hjust=.5, color="navy"),
  panel.grid.minor=element_blank(),
  axis.text.x = element_text(angle=45, hjust=1)
)

theme_set(SamTheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The lake temperature does NOT change with depth across all lakes
Ha: The lake temperature does change with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.



```{r scatterplot, fig.cap = "Temperature By Depth", fig.height=3.5}
#4

NTLakesWrangled<-NTLakes %>% 
  filter(month(sampledate) == 7) %>% 
  select(lakename:daynum, depth:temperature_C) %>% 
  drop_na(lakename:daynum, depth:temperature_C)


#5

NTLakesScatter<-ggplot(NTLakesWrangled,
                        aes(x=depth,
                          y=temperature_C, 
                          color=temperature_C))+
                        geom_point(alpha=.6)+
                        scale_color_gradient(low= 'purple', 
                          high= 'orange')+
                          ylim(0, 35)+
                            geom_smooth(method="lm", 
                                        color='green')+
                            labs(color= "Temperature", 
                                 x= 'Depth (meter)', 
                                 y= 'Temperature (Celsius)', 
                                 title= 'Temperature by Depth'
                              )+
  theme(plot.title=element_text(hjust=.5, color='navy'),
        legend.position = 'none')

print(NTLakesScatter)
       


```

\newpage
6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: I think the graph clearly shows an an inverse relationship between temperature and depth. As the depth increases, the temperature decreases. However, I think lm is not the best representation of this relationship. It does look like as the depth approaches around 10 meters, the temperature stops decreases much past 5 degrees. I think a more curved regression how show the changing nature of this relationship.


7. Perform a linear regression to test the relationship and display the results.



```{r linear.regression}
#7

TempDepthLM<-lm(data=NTLakesWrangled, temperature_C ~ depth)
summary(TempDepthLM)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression shows a statistically-signifgant inverse
relationshio between temperature and depth. The depth coefficient is -1.94
which means for every meter depth increases, temperature will decrease by about
2 degrees. Additionally, when depth is 0, its predicted that the temperature
will be 21.95 degrees. This is significant beyond the.001 point, way below the
.05 threshold we tend to use, so we know that these results are statistically
significant. The multiple r-squared is .738 which also indicates that ~73%
of the variance of temperature is explained by depth, a high figure

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

TempDepthAIC<-lm(data=NTLakesWrangled, temperature_C ~ year4 + daynum + depth)

step(TempDepthAIC)



#10

#initial lm TempDepthAIC was optimal


TempDepthRegression<-lm(data=NTLakesWrangled, temperature_C ~ year4 + daynum + depth)

summary(TempDepthRegression)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 

The final set of explanatory variable to predict temperature according
to the AIC model is year, day, and depth, with removing none being the optimal
AIC. A linear regression of these three variables leads to an r squared
value of .74, meaning these three variables explain 74% of the variance of
temperatrure, rather than the 73% explained by just depth. That means that the
multiple regression is a slightly better fitted
model, even if they are both high.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

NTLakesTempJulyANOVA<-lm(data=NTLakesWrangled, temperature_C ~ lakename)
anova(NTLakesTempJulyANOVA)

NTLakesTempJulyAOV<-aov(data=NTLakesWrangled, temperature_C ~ lakename)
summary(NTLakesTempJulyAOV)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 

Yes, the P value for both of them says that the result is signifcgant, therefore
confirming the alternative hypothesis and saying we know that based on 
the sample, we can assume the mean of the the different lakes are different.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2, fig.cap= 'Temperature by Depth per Lake'}
#14.

TempByDepthPlot<-ggplot(NTLakesWrangled,aes(x=depth, 
                                             y=temperature_C, 
                                             color=lakename))+
                          geom_point(alpha=.5)+
                          ylim(0,35)+
                          geom_smooth(method='lm', se=F)+
                          labs(y='Temperature (Celsius)', 
                               x='Depth (Meters)',
                               color='Lake Name',
                               title= 'Temperature by Depth Per Lake')
                            plot.title=element_text(hjust=.5)
                          
                          
print(TempByDepthPlot)


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15

TukeyHSD(NTLakesTempJulyAOV)


NTLakesHSD<-HSD.test(NTLakesTempJulyAOV, 'lakename', group=TRUE)

print(NTLakesHSD)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 Same as Peter Lake: Paul Lake, Ward Lake
 Distinct from all others: None of the, all of are similar to at least one other

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer:

We can use a t-test, two-sample test is used to test the hypothesis that the mean of two samples is equivalent. Unlike the one-sample tests, a two-sample test requires a second assumption that the variance of the two groups is equivalent. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

NTLakesTTestData<-NTLakesWrangled %>% 
  filter(NTLakesWrangled$lakename== c("Crampton Lake", "Ward Lake"))

view(NTLakesTTestData)

NTLakesTTest<-t.test(NTLakesTTestData$temperature_C ~ NTLakesTTestData$lakename)
NTLakesTTest

```

>Answer: The test says the mean temp of Ward and Crampton lake are not identical
but similar, and they match the results of the Tukey Test within .1 of a degree.
Based on the Tukey test, even though these means are not identical they are
statistically speaking, as they fall within the range of each other.
